---
title: "The 'ratioOfQsprays' package"
author: "Stéphane Laurent"
date: "`r Sys.Date()`"
output: github_document
---

*Fractions of multivariate polynomials with rational coefficients.*

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/ratioOfQsprays/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/ratioOfQsprays/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

___

The **qspray** package allows arithmetic (and more) on multivariate polynomials 
with rational coefficients. Based on this one, the **ratioOfQsprays** package 
allows to manipulate fractions of multivariate polynomials with rational 
coefficients.

## Creating a `ratioOfQsprays` 

A `ratioOfQsprays` object represents a fraction of two multivariate 
polynomial with rational coefficients. Such polynomials are represented by 
`qspray` objects. The easiest way to create a `ratioOfQsprays` is to introduce 
the variables of the polynomials with the `qlone` function, and then to build a 
`qspray` numerator and a `qspray` denominator with arithmetic operations. 
For example:

```{r, message=FALSE, collapse=TRUE}
library(ratioOfQsprays)
f <- function(x1, x2, x3) {
  (2*x1^2 + x2*x3) / (4*x1 - 3*x3 + 1)
}
# variables:
x1 <- qlone(1)
x2 <- qlone(2)
x3 <- qlone(3)
# the 'ratioOfQsprays':
( roq <- f(x1, x2, x3) )
```

Arithmetic on `ratioOfQsprays` objects is available:

```{r, message=FALSE, collapse=TRUE}
roq^2
roq - roq
1 / roq
roq + (x2 + x3)/x1
```

Rational numbers and \code{qspray} polynomials are coercable to 
`ratioOfQsprays` objects, and you can also perform arithmetic operations 
between a `ratioOfQsprays` and such an object:

```{r, message=FALSE, collapse=TRUE}
2 * roq
"1/2" * roq
roq + gmp::as.bigq("7/3") 
x1 + roq + x3
```

## Evaluating a `ratioOfQsprays` 

Use `evalRatioOfQsprays` to evaluate a `ratioOfQsprays`. This function returns 
a `bigq` number:

```{r, message=FALSE, collapse=TRUE}
library(gmp) # rational numbers
x <- c("4", "3", "2/5")
evalRatioOfQsprays(roq, x)
x <- as.bigq(x)
evalRatioOfQsprays(roq, x)
f(x[1], x[2], x[3])
```
