---
title: "The 'ratioOfQsprays' package"
author: "Stéphane Laurent"
date: "`r Sys.Date()`"
output: github_document
---

*Fractions of multivariate polynomials with rational coefficients.*

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/ratioOfQsprays/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/ratioOfQsprays/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

___

The **qspray** package allows arithmetic (and more) on multivariate polynomials 
with rational coefficients. Based on this one, the **ratioOfQsprays** package 
allows to manipulate fractions of multivariate polynomials with rational 
coefficients.

A `symbolicQspray` object represents a multivariate polynomial whose 
coefficients are fractions of polynomials. For example:

```{r, message=FALSE, collapse=TRUE}
library(ratioOfQsprays)
f <- function(x1, x2, x3) {
  2*x1^2 / (3*x3 + 1)  +  (x2 * x3) / (4*x1 + x2)
}
# variables:
x1 <- qlone(1)
x2 <- qlone(2)
x3 <- qlone(3)
# the 'ratioOfQsprays':
( roq <- f(x1, x2, x3) )
```

Arithmetic on `ratioOfQsprays` objects is available:

```{r, message=FALSE, collapse=TRUE}
roq^2
roq - roq
1 / roq
```

And evaluation: 

```{r, message=FALSE, collapse=TRUE}
library(gmp) # rational numbers
x <- c(as.bigq(4), as.bigq(3), as.bigq("2/5"))
evalRatioOfQsprays(roq, x)
f(x[1], x[2], x[3])
```
